---
title: "Data Overview & Exploration"
subtitle: "TCR Repertoire Analysis — Notebook 01"
author: "Joshua Luthy"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: null
    css: "../www/report-style.css"
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(tidyverse)
```

## Dataset Overview

This notebook provides initial exploration of the synthetic TCR repertoire dataset.
The data simulates a **cell therapy manufacturing workflow** where PBMCs from the
**apheresis** collection are processed into a final **drug product**.

```{r load-data}
# Load synthetic data
tcr_data <- read_csv("../data/raw/synthetic_tcr_repertoires.csv")
patient_meta <- read_csv("../data/raw/patient_metadata.csv")

glimpse(tcr_data)
```

## Patient Cohort

```{r patient-table}
patient_meta %>%
  knitr::kable(caption = "Patient cohort summary")
```

## Per-Sample Summary

```{r diversity-summary}
diversity_summary <- tcr_data %>%
  group_by(patient_id, sample_type, clinical_response) %>%
  summarise(
    n_clonotypes    = n_distinct(clonotype_id),
    total_reads     = sum(clone_count),
    shannon_entropy = vegan::diversity(clone_count, index = "shannon"),
    clonality       = 1 - (shannon_entropy / log(n_clonotypes)),
    .groups = "drop"
  )

diversity_summary %>%
  knitr::kable(digits = 4, caption = "Diversity metrics per sample")
```

## V Gene Usage

```{r vgene-usage, fig.height=8}
vgene_summary <- tcr_data %>%
  group_by(sample_type, v_gene) %>%
  summarise(total_count = sum(clone_count), .groups = "drop") %>%
  group_by(sample_type) %>%
  mutate(fraction = total_count / sum(total_count)) %>%
  slice_max(fraction, n = 10)

ggplot(vgene_summary, aes(x = reorder(v_gene, fraction), y = fraction,
                           fill = sample_type)) +
  geom_col(position = "dodge", alpha = 0.8) +
  coord_flip() +
  scale_fill_manual(values = c("Apheresis" = "#00d4ff", "Product" = "#7b2fff")) +
  labs(x = NULL, y = "Read Fraction", fill = "Sample Type",
       title = "Top 10 TRBV Gene Segments") +
  theme_minimal()
```

## Data Quality Notes

> **Synthetic Data Disclaimer:** This dataset is entirely synthetic, generated to
> demonstrate analytical methods. CDR3 sequences are randomly generated and do not
> correspond to real TCR rearrangements.
