---
title: "Data Overview & Exploration"
subtitle: "TCR Repertoire Analysis — Notebook 01"
author: "Joshua Luthy"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    css: "../www/report-style.css"
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 6,
  dev = "png", dpi = 150,
  fig.retina = 2
)
library(tidyverse)

# Dark theme for all ggplot output — matches portfolio site
theme_tcr <- function(base_size = 13) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.background    = element_rect(fill = "#0a0f14", color = NA),
      panel.background   = element_rect(fill = "#0a0f14", color = NA),
      panel.grid.major   = element_line(color = "#1e3348", linewidth = 0.3),
      panel.grid.minor   = element_blank(),
      text               = element_text(color = "#e0eef5", family = "sans"),
      axis.text          = element_text(color = "#7b9db5", size = 10),
      axis.title         = element_text(color = "#7b9db5", size = 11),
      plot.title         = element_text(color = "#e0eef5", face = "bold", size = 16,
                                        margin = margin(b = 8)),
      plot.subtitle      = element_text(color = "#7b9db5", size = 11,
                                        margin = margin(b = 12)),
      legend.background  = element_rect(fill = "#0f1923", color = "#1e3348"),
      legend.text        = element_text(color = "#7b9db5"),
      legend.title       = element_text(color = "#e0eef5"),
      strip.text         = element_text(color = "#e0eef5", face = "bold"),
      plot.margin        = margin(15, 15, 15, 15)
    )
}

theme_set(theme_tcr())
```

## Dataset Overview

This notebook provides initial exploration of the synthetic TCR repertoire dataset.
The data simulates a **cell therapy manufacturing workflow** where PBMCs from the
**apheresis** collection are processed into a final **drug product**.

```{r load-data}
tcr_data <- read_csv("../data/raw/synthetic_tcr_repertoires.csv")
patient_meta <- read_csv("../data/raw/patient_metadata.csv")

n_patients    <- n_distinct(tcr_data$patient_id)
n_samples     <- n_distinct(paste(tcr_data$patient_id, tcr_data$sample_type))
n_clonotypes  <- nrow(tcr_data)
n_reads       <- sum(tcr_data$clone_count)
```

```{r metrics-cards, echo=FALSE}
htmltools::div(class = "metrics-row",
  htmltools::div(class = "metric-card",
    htmltools::div(class = "value", n_patients),
    htmltools::div(class = "label", "Patients")
  ),
  htmltools::div(class = "metric-card",
    htmltools::div(class = "value", n_samples),
    htmltools::div(class = "label", "Samples")
  ),
  htmltools::div(class = "metric-card",
    htmltools::div(class = "value", format(n_clonotypes, big.mark = ",")),
    htmltools::div(class = "label", "Total Clonotypes")
  ),
  htmltools::div(class = "metric-card",
    htmltools::div(class = "value", format(n_reads, big.mark = ",")),
    htmltools::div(class = "label", "Total Reads")
  )
)
```

```{r glimpse}
glimpse(tcr_data)
```

## Patient Cohort

The cohort includes **`r n_patients` AML patients** with varying clinical responses
to cell therapy: 2 complete responders (CR), 2 partial responders (PR), and
2 with progressive disease (PD).

```{r patient-table}
patient_meta %>%
  knitr::kable(caption = "Patient cohort summary")
```

## Per-Sample Summary

For each patient, we have paired **Apheresis** (starting material) and **Product**
(manufactured drug product) TCR repertoire samples.

```{r diversity-summary}
# Custom Shannon entropy (avoids vegan dependency for portability)
shannon_entropy <- function(counts) {
  p <- counts / sum(counts)
  p <- p[p > 0]
  -sum(p * log(p))
}

diversity_summary <- tcr_data %>%
  group_by(patient_id, sample_type, clinical_response) %>%
  summarise(
    n_clonotypes    = n_distinct(clonotype_id),
    total_reads     = sum(clone_count),
    shannon_entropy = shannon_entropy(clone_count),
    clonality       = 1 - (shannon_entropy / log(n_clonotypes)),
    .groups = "drop"
  )

diversity_summary %>%
  mutate(across(c(shannon_entropy, clonality), ~round(., 4))) %>%
  knitr::kable(
    caption = "Diversity metrics per sample",
    format.args = list(big.mark = ",")
  )
```

> **Key Observation:** Apheresis samples show **high diversity** (Shannon H > 7.0,
> low clonality) with thousands of unique clonotypes, while Product samples are
> dramatically more **oligoclonal** (Shannon H 3.0–5.3, clonality 0.34–0.55).
> This is consistent with the expected biology: manufacturing selects and expands
> a subset of T cell clones from the starting material.

## V Gene Usage

We examine TRBV gene segment usage across samples to identify preferential
gene usage patterns between Apheresis and Product.

```{r vgene-usage, fig.height=7}
vgene_summary <- tcr_data %>%
  group_by(sample_type, v_gene) %>%
  summarise(total_count = sum(clone_count), .groups = "drop") %>%
  group_by(sample_type) %>%
  mutate(fraction = total_count / sum(total_count)) %>%
  slice_max(fraction, n = 10) %>%
  ungroup()

ggplot(vgene_summary, aes(x = reorder(v_gene, fraction), y = fraction,
                           fill = sample_type)) +
  geom_col(position = "dodge", alpha = 0.8, width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("Apheresis" = "#00d4ff", "Product" = "#7b2fff")) +
  labs(x = NULL, y = "Read Fraction", fill = "Sample Type",
       title = "Top 10 TRBV Gene Segments",
       subtitle = "Read fraction by sample type, aggregated across all patients")
```

## Data Quality Notes

> **Synthetic Data Disclaimer:** This dataset is entirely synthetic, generated to
> demonstrate analytical methods and pipeline architecture. CDR3 sequences are
> randomly generated and do not correspond to real TCR rearrangements. Clone size
> distributions follow power-law models calibrated to approximate real repertoire
> characteristics.

Data quality checks:

- All CDR3 sequences begin with C and end with F (conserved anchors) ✓
- Clone fractions sum to ~1.0 within each sample ✓
- No duplicate clonotype IDs within samples ✓
- All patients have paired Apheresis + Product samples ✓
- TRBV/TRBJ gene names match IMGT nomenclature ✓
