---
title: "Clonal Expansion & Dominance Analysis"
subtitle: "TCR Repertoire Analysis — Notebook 03"
author: "Joshua Luthy"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: null
    css: "../www/report-style.css"
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(tidyverse)
```

## Analysis Overview

This notebook examines **clonal expansion patterns** in the manufactured Product,
identifying the most dominant clonotypes and characterizing oligoclonality.

```{r load-data}
tcr_data <- read_csv("../data/raw/synthetic_tcr_repertoires.csv")
top_clones <- read_csv("../data/processed/top_clonotypes.csv")
```

## Top Expanded Clonotypes

```{r top-clones-table}
top_clones %>%
  filter(patient_id == "PT-001", sample_type == "Product") %>%
  head(10) %>%
  mutate(clone_fraction = scales::percent(clone_fraction, accuracy = 0.01)) %>%
  knitr::kable(caption = "Top 10 expanded clonotypes — PT-001 (CR)")
```

## Clonal Dominance Patterns

```{r cumulative-plot}
cumulative <- top_clones %>%
  filter(sample_type == "Product") %>%
  group_by(patient_id) %>%
  arrange(rank) %>%
  mutate(cum_fraction = cumsum(clone_fraction))

ggplot(cumulative, aes(x = rank, y = cum_fraction,
                        color = clinical_response, group = patient_id)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  labs(x = "Clonotype Rank", y = "Cumulative Fraction",
       color = "Response", title = "Cumulative Clonal Dominance in Product") +
  theme_minimal()
```

## Shared Clonotype Analysis

```{r shared-analysis}
# Identify shared clonotypes between Apheresis and Product
aph_clones <- tcr_data %>% filter(sample_type == "Apheresis") %>%
  select(patient_id, cdr3_aa) %>% distinct()

prod_clones <- tcr_data %>% filter(sample_type == "Product") %>%
  select(patient_id, cdr3_aa) %>% distinct()

shared <- inner_join(aph_clones, prod_clones, by = c("patient_id", "cdr3_aa"))

overlap_summary <- prod_clones %>%
  group_by(patient_id) %>%
  summarise(
    total_product = n(),
    shared = sum(cdr3_aa %in% shared$cdr3_aa[shared$patient_id == patient_id[1]]),
    pct_shared = shared / total_product * 100
  )

overlap_summary %>% knitr::kable(digits = 1, caption = "Apheresis-Product clonotype overlap")
```

## Conclusions

1. Manufacturing drives extreme oligoclonal expansion.
2. CR patients show the most concentrated Product repertoires.
3. The majority of dominant Product clones are traceable to the Apheresis starting material.
4. PD patients exhibit more diffuse Product repertoires.
