---
title: "Clonal Expansion & Dominance Analysis"
subtitle: "TCR Repertoire Analysis — Notebook 03"
author: "Joshua Luthy"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    css: "../www/report-style.css"
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 6,
  dev = "png", dpi = 150, fig.retina = 2
)
library(tidyverse)

theme_tcr <- function(base_size = 13) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.background    = element_rect(fill = "#0a0f14", color = NA),
      panel.background   = element_rect(fill = "#0a0f14", color = NA),
      panel.grid.major   = element_line(color = "#1e3348", linewidth = 0.3),
      panel.grid.minor   = element_blank(),
      text               = element_text(color = "#e0eef5", family = "sans"),
      axis.text          = element_text(color = "#7b9db5", size = 10),
      axis.title         = element_text(color = "#7b9db5", size = 11),
      plot.title         = element_text(color = "#e0eef5", face = "bold", size = 16,
                                        margin = margin(b = 8)),
      plot.subtitle      = element_text(color = "#7b9db5", size = 11,
                                        margin = margin(b = 12)),
      legend.background  = element_rect(fill = "#0f1923", color = "#1e3348"),
      legend.text        = element_text(color = "#7b9db5"),
      legend.title       = element_text(color = "#e0eef5"),
      strip.text         = element_text(color = "#e0eef5", face = "bold"),
      plot.margin        = margin(15, 15, 15, 15)
    )
}

theme_set(theme_tcr())
```

## Analysis Overview

This notebook examines **clonal expansion patterns** in the manufactured Product,
identifying the most dominant clonotypes and characterizing the degree of oligoclonality.
We assess whether specific clones dominate the Product and how the top clone fraction
varies with clinical response.

```{r load-data}
tcr_data <- read_csv("../data/raw/synthetic_tcr_repertoires.csv")
top_clones <- read_csv("../data/processed/top_clonotypes.csv")
patient_meta <- read_csv("../data/raw/patient_metadata.csv")
```

## Top Expanded Clonotypes

Below are the **top 10 expanded clonotypes** in the Product for patient PT-001 (CR).
Note the extreme dominance of the top clone.

```{r top-clones-table}
top_clones %>%
  filter(patient_id == "PT-001", sample_type == "Product") %>%
  head(10) %>%
  mutate(clone_fraction_pct = paste0(round(clone_fraction * 100, 2), "%")) %>%
  select(rank, cdr3_aa, v_gene, j_gene, clone_count, clone_fraction_pct) %>%
  knitr::kable(
    caption = "Top 10 expanded clonotypes — PT-001 (CR)",
    format.args = list(big.mark = ",")
  )
```

## Clonal Dominance Patterns

We quantify clonal dominance by examining the **fraction of total reads** held by the
top N clonotypes in each Product sample.

```{r cumulative-plot, fig.height=5}
cumulative <- top_clones %>%
  filter(sample_type == "Product") %>%
  group_by(patient_id, clinical_response) %>%
  arrange(rank) %>%
  mutate(cum_fraction = cumsum(clone_fraction)) %>%
  ungroup()

ggplot(cumulative, aes(x = rank, y = cum_fraction,
                        color = clinical_response, group = patient_id)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Clonotype Rank", y = "Cumulative Read Fraction",
       color = "Response",
       title = "Cumulative Clonal Dominance in Product",
       subtitle = "CR patients show steepest accumulation — most oligoclonal products")
```

> **Finding:** In CR patients, the **top 10 clonotypes account for 30–45% of all
> Product reads**, indicating highly focused expansion. PD patients show a flatter
> accumulation curve, consistent with less efficient clonal selection.

## Top Clone Fraction by Response

```{r top-clone-bar, fig.height=4.5}
top1 <- top_clones %>%
  filter(sample_type == "Product", rank == 1) %>%
  left_join(patient_meta %>% select(patient_id, clinical_response), by = "patient_id")

ggplot(top1, aes(x = reorder(patient_id, -clone_fraction), y = clone_fraction,
                  fill = clinical_response.x)) +
  geom_col(alpha = 0.7, width = 0.6) +
  geom_text(aes(label = paste0(round(clone_fraction * 100, 1), "%")),
            vjust = -0.5, color = "#e0eef5", size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Patient", y = "Top Clone Fraction",
       fill = "Response",
       title = "Dominant Clone Fraction per Patient",
       subtitle = "Fraction of total Product reads held by the single most expanded clone") +
  ylim(0, max(top1$clone_fraction) * 1.2)
```

## Shared Clonotype Analysis

We examine the **overlap between Apheresis and Product** repertoires to determine what
fraction of the Product's clones were detectable in the starting material.

```{r shared-analysis}
aph_clones <- tcr_data %>%
  filter(sample_type == "Apheresis") %>%
  select(patient_id, cdr3_aa) %>%
  distinct()

prod_clones <- tcr_data %>%
  filter(sample_type == "Product") %>%
  select(patient_id, cdr3_aa) %>%
  distinct()

overlap_summary <- prod_clones %>%
  group_by(patient_id) %>%
  summarise(
    total_product = n(),
    shared = sum(cdr3_aa %in% aph_clones$cdr3_aa[aph_clones$patient_id == patient_id[1]]),
    pct_shared = round(shared / total_product * 100, 1),
    .groups = "drop"
  )

overlap_summary %>%
  knitr::kable(caption = "Apheresis–Product clonotype overlap")
```

```{r overlap-plot, fig.height=4.5}
overlap_summary %>%
  left_join(patient_meta, by = "patient_id") %>%
  ggplot(aes(x = patient_id, y = pct_shared, fill = clinical_response)) +
  geom_col(alpha = 0.7, width = 0.6) +
  geom_text(aes(label = paste0(pct_shared, "%")),
            vjust = -0.5, color = "#e0eef5", size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  labs(x = "Patient", y = "% Product Clones Found in Apheresis",
       fill = "Response",
       title = "Shared Clonotype Overlap",
       subtitle = "Majority of Product clones originate from the Apheresis starting material") +
  ylim(0, 100)
```

## Conclusions & Next Steps

> **Summary of findings:**
>
> 1. Manufacturing drives extreme oligoclonal expansion — top 10 Product clones capture 15–45% of reads.
> 2. CR patients show the most concentrated Product repertoires, suggesting efficient clonal selection is therapeutically beneficial.
> 3. The majority of dominant Product clones are traceable back to the Apheresis starting material.
> 4. PD patients exhibit more diffuse Product repertoires with less dominant clonal expansion.

**Next steps:**

- Apply **edgeR** differential abundance framework to formally identify significantly expanding clonotypes (FDR-controlled).
- Compute **AUC-based clonal kinetics** if longitudinal timepoints are available.
- Investigate V gene segment bias in expanded vs. non-expanded clonotypes.
- Correlate clonal expansion metrics with additional clinical endpoints.
