---
title: "edgeR Differential Abundance Analysis"
subtitle: "TCR Repertoire Analysis — Notebook 04"
author: "Joshua Luthy"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    css: "../www/report-style.css"
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 6,
  dev = "png", dpi = 150, fig.retina = 2
)
library(tidyverse)
library(edgeR)

theme_tcr <- function(base_size = 13) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.background    = element_rect(fill = "#0a0f14", color = NA),
      panel.background   = element_rect(fill = "#0a0f14", color = NA),
      panel.grid.major   = element_line(color = "#1e3348", linewidth = 0.3),
      panel.grid.minor   = element_blank(),
      text               = element_text(color = "#e0eef5", family = "sans"),
      axis.text          = element_text(color = "#7b9db5", size = 10),
      axis.title         = element_text(color = "#7b9db5", size = 11),
      plot.title         = element_text(color = "#e0eef5", face = "bold", size = 16,
                                        margin = margin(b = 8)),
      plot.subtitle      = element_text(color = "#7b9db5", size = 11,
                                        margin = margin(b = 12)),
      legend.background  = element_rect(fill = "#0f1923", color = "#1e3348"),
      legend.text        = element_text(color = "#7b9db5"),
      legend.title       = element_text(color = "#e0eef5"),
      strip.text         = element_text(color = "#e0eef5", face = "bold"),
      plot.margin        = margin(15, 15, 15, 15)
    )
}

theme_set(theme_tcr())
```

## Analysis Overview

This notebook applies the **edgeR differential abundance framework** to identify
individual clonotypes that significantly **expand** or **contract** from Apheresis to
Product. Rather than summarizing diversity at the sample level (as in Notebook 02),
here we test each clonotype for differential abundance using edgeR's negative binomial
generalized linear model (GLM) with quasi-likelihood F-tests and FDR correction.

This approach is adapted from differential gene expression analysis and treats each
clonotype analogously to a gene, with clone counts as the "expression" measure across
conditions (Apheresis vs Product).

```{r load-data}
tcr_data <- read_csv("../data/raw/synthetic_tcr_repertoires.csv")
patient_meta <- read_csv("../data/raw/patient_metadata.csv")

# We will analyze each patient separately to identify patient-specific
# expanding and contracting clonotypes, then look for patterns across patients.
```

## Preparing the Count Matrix

For edgeR, we need a **count matrix** where rows are clonotypes (identified by CDR3
sequence) and columns are samples. We focus on clonotypes observed in at least one
condition per patient and build a paired Apheresis vs Product design.

```{r build-count-matrix}
# Function to build a count matrix for one patient
build_patient_counts <- function(df, pid) {
  patient_df <- df %>% filter(patient_id == pid)
  
  # Pivot to wide: rows = CDR3, columns = sample_type
  count_wide <- patient_df %>%
    select(cdr3_aa, sample_type, clone_count) %>%
    pivot_wider(
      names_from  = sample_type,
      values_from = clone_count,
      values_fill = 0
    )
  
  # Convert to matrix
  count_mat <- count_wide %>%
    select(Apheresis, Product) %>%
    as.matrix()
  rownames(count_mat) <- count_wide$cdr3_aa
  
  return(count_mat)
}

# Example: PT-001
count_mat_pt001 <- build_patient_counts(tcr_data, "PT-001")
dim(count_mat_pt001)
```

```{r count-preview}
# Preview the top clonotypes by Product count
head(count_mat_pt001[order(-count_mat_pt001[, "Product"]), ], 10)
```

## edgeR Model Setup

We use edgeR's **quasi-likelihood (QL) framework**, which provides more robust
inference for small sample sizes. The design tests for differences between
Apheresis and Product within each patient.

```{r edger-single-patient}
# Function to run edgeR for a single patient
run_edger_patient <- function(count_mat, patient_id) {
  # Create DGEList
  group <- factor(colnames(count_mat), levels = c("Apheresis", "Product"))
  y <- DGEList(counts = count_mat, group = group)
  
  # Filter low-count clonotypes (must have >= 5 counts in at least one sample)
  keep <- rowSums(count_mat >= 5) >= 1
  y <- y[keep, , keep.lib.sizes = FALSE]
  
  # Normalization (TMM)
  y <- calcNormFactors(y)
  
  # Design matrix
  design <- model.matrix(~ group)
  
  # Dispersion estimation
  # With only 2 samples per patient (no biological replicates), edgeR cannot
  # estimate dispersion from the data. We set a fixed biological coefficient
  # of variation (BCV). A BCV of 0.4 (dispersion = 0.16) is a reasonable
  # assumption for immune repertoire data, consistent with recommendations
  # in the edgeR user guide for unreplicated designs.
  bcv <- 0.4
  y$common.dispersion <- bcv^2
  
  # Fit GLM with fixed dispersion
  fit <- glmFit(y, design, dispersion = bcv^2)
  
  # Likelihood ratio test for Product vs Apheresis
  lrt <- glmLRT(fit, coef = 2)
  
  # Extract results
  results <- topTags(lrt, n = Inf, sort.by = "PValue")$table %>%
    as_tibble(rownames = "cdr3_aa") %>%
    mutate(
      patient_id = patient_id,
      significant = FDR < 0.05,
      direction = case_when(
        significant & logFC > 0 ~ "Expander",
        significant & logFC < 0 ~ "Contractor",
        TRUE ~ "NS"
      )
    )
  
  return(results)
}

# Run for PT-001
results_pt001 <- run_edger_patient(count_mat_pt001, "PT-001")

# Summary
results_pt001 %>%
  count(direction) %>%
  knitr::kable(caption = "PT-001: Differential abundance summary (FDR < 0.05)")
```

## Running Across All Patients

```{r run-all-patients}
# Run edgeR for all patients
all_results <- map_dfr(unique(tcr_data$patient_id), function(pid) {
  count_mat <- build_patient_counts(tcr_data, pid)
  run_edger_patient(count_mat, pid)
})

# Join clinical response
all_results <- all_results %>%
  left_join(patient_meta %>% select(patient_id, clinical_response), by = "patient_id")

# Overall summary
summary_table <- all_results %>%
  group_by(patient_id, clinical_response, direction) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = direction, values_from = n, values_fill = 0)

summary_table %>%
  knitr::kable(caption = "Differential abundance summary across all patients")
```

```{r summary-metrics, echo=FALSE}
total_expanders   <- sum(all_results$direction == "Expander")
total_contractors <- sum(all_results$direction == "Contractor")
total_tested      <- nrow(all_results)
pct_sig           <- round((total_expanders + total_contractors) / total_tested * 100, 1)

htmltools::div(class = "metrics-row",
  htmltools::div(class = "metric-card",
    htmltools::div(class = "value", format(total_tested, big.mark = ",")),
    htmltools::div(class = "label", "Clonotypes Tested")
  ),
  htmltools::div(class = "metric-card",
    htmltools::div(class = "value", style = "color:#00ff9d", format(total_expanders, big.mark = ",")),
    htmltools::div(class = "label", "Significant Expanders")
  ),
  htmltools::div(class = "metric-card",
    htmltools::div(class = "value", style = "color:#ff6b6b", format(total_contractors, big.mark = ",")),
    htmltools::div(class = "label", "Significant Contractors")
  ),
  htmltools::div(class = "metric-card",
    htmltools::div(class = "value", style = "color:#ffd93d", paste0(pct_sig, "%")),
    htmltools::div(class = "label", "% Significant (FDR < 0.05)")
  )
)
```

## Volcano Plots

Volcano plots display the **log₂ fold change** (x-axis) against the **−log₁₀ FDR**
(y-axis) for each clonotype. Points above the significance threshold (dashed line)
and beyond the fold-change cutoffs are colored as expanders (green) or contractors (red).

### Individual Patient Volcano Plots

```{r volcano-individual, fig.height=10, fig.width=12}
# Prepare volcano data with capped values for visualization
volcano_data <- all_results %>%
  mutate(
    neg_log10_fdr = -log10(pmax(FDR, 1e-50)),  # cap for visualization
    logFC_capped  = pmax(pmin(logFC, 15), -15),  # cap extreme FCs
    direction     = factor(direction, levels = c("Contractor", "NS", "Expander"))
  )

ggplot(volcano_data, aes(x = logFC_capped, y = neg_log10_fdr, color = direction)) +
  geom_point(alpha = 0.4, size = 1.2) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "#ffd93d", linewidth = 0.5) +
  geom_vline(xintercept = c(-1, 1), linetype = "dotted", color = "#3f6380", linewidth = 0.4) +
  scale_color_manual(
    values = c("Contractor" = "#ff6b6b", "NS" = "#3f6380", "Expander" = "#00ff9d"),
    labels = c("Contractor", "Not Significant", "Expander")
  ) +
  facet_wrap(~ patient_id, scales = "free", ncol = 3) +
  labs(
    x = expression(log[2] ~ "Fold Change (Product / Apheresis)"),
    y = expression(-log[10] ~ "FDR"),
    color = "Status",
    title = "Volcano Plots: Clonotype Differential Abundance",
    subtitle = "Each panel shows one patient — dashed line: FDR = 0.05, dotted lines: |log₂FC| = 1"
  ) +
  theme(
    strip.background = element_rect(fill = "#152231", color = "#1e3348"),
    strip.text       = element_text(color = "#00d4ff", face = "bold", size = 11),
    legend.position  = "bottom"
  )
```

### Combined Volcano Plot (All Patients)

```{r volcano-combined, fig.height=6}
ggplot(volcano_data, aes(x = logFC_capped, y = neg_log10_fdr, color = direction)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "#ffd93d", linewidth = 0.6) +
  geom_vline(xintercept = c(-1, 1), linetype = "dotted", color = "#3f6380", linewidth = 0.4) +
  scale_color_manual(
    values = c("Contractor" = "#ff6b6b", "NS" = "#3f6380", "Expander" = "#00ff9d"),
    labels = c("Contractor", "Not Significant", "Expander")
  ) +
  labs(
    x = expression(log[2] ~ "Fold Change (Product / Apheresis)"),
    y = expression(-log[10] ~ "FDR"),
    color = "Status",
    title = "Combined Volcano Plot — All Patients",
    subtitle = "Clonotypes significantly expanding (green) or contracting (red) from Apheresis to Product"
  ) +
  theme(legend.position = "bottom")
```

> **Interpretation:** The volcano plots reveal a clear asymmetry — many more clonotypes
> are significantly **contracting** (lost or reduced in Product) than expanding. This is
> expected: manufacturing selects a small subset of clones for expansion while the vast
> majority of apheresis-origin clonotypes are depleted.

## Top Expanders by Patient

```{r top-expanders}
top_expanders <- all_results %>%
  filter(direction == "Expander") %>%
  group_by(patient_id) %>%
  slice_max(logFC, n = 5) %>%
  ungroup() %>%
  select(patient_id, clinical_response, cdr3_aa, logFC, logCPM, FDR) %>%
  mutate(
    logFC = round(logFC, 2),
    logCPM = round(logCPM, 2),
    FDR = formatC(FDR, format = "e", digits = 2)
  )

top_expanders %>%
  knitr::kable(caption = "Top 5 significantly expanding clonotypes per patient (by log₂FC)")
```

## Top Contractors by Patient

```{r top-contractors}
top_contractors <- all_results %>%
  filter(direction == "Contractor") %>%
  group_by(patient_id) %>%
  slice_min(logFC, n = 5) %>%
  ungroup() %>%
  select(patient_id, clinical_response, cdr3_aa, logFC, logCPM, FDR) %>%
  mutate(
    logFC = round(logFC, 2),
    logCPM = round(logCPM, 2),
    FDR = formatC(FDR, format = "e", digits = 2)
  )

top_contractors %>%
  knitr::kable(caption = "Top 5 significantly contracting clonotypes per patient (by log₂FC)")
```

## Expander/Contractor Counts by Clinical Response

```{r direction-by-response, fig.height=5}
direction_counts <- all_results %>%
  filter(direction != "NS") %>%
  group_by(patient_id, clinical_response, direction) %>%
  summarise(n = n(), .groups = "drop")

ggplot(direction_counts, aes(x = patient_id, y = n, fill = direction)) +
  geom_col(position = "dodge", alpha = 0.8, width = 0.6) +
  scale_fill_manual(values = c("Expander" = "#00ff9d", "Contractor" = "#ff6b6b")) +
  labs(
    x = "Patient", y = "Number of Clonotypes",
    fill = "Direction",
    title = "Significant Expanders vs Contractors per Patient",
    subtitle = "FDR < 0.05 — colored by patient clinical response group"
  ) +
  geom_text(aes(label = n), position = position_dodge(width = 0.6),
            vjust = -0.5, color = "#e0eef5", size = 3.2) +
  facet_grid(~ clinical_response, scales = "free_x", space = "free_x") +
  theme(
    strip.background = element_rect(fill = "#152231", color = "#1e3348"),
    strip.text       = element_text(color = "#00d4ff", face = "bold", size = 11)
  )
```

## Fold Change Distribution

```{r fc-distribution, fig.height=5}
all_results %>%
  filter(significant) %>%
  ggplot(aes(x = logFC, fill = direction)) +
  geom_histogram(bins = 60, alpha = 0.7, color = NA) +
  scale_fill_manual(values = c("Expander" = "#00ff9d", "Contractor" = "#ff6b6b", "NS" = "#3f6380")) +
  geom_vline(xintercept = 0, linetype = "solid", color = "#7b9db5", linewidth = 0.5) +
  labs(
    x = expression(log[2] ~ "Fold Change"),
    y = "Count",
    fill = "Direction",
    title = "Distribution of Significant log₂ Fold Changes",
    subtitle = "Significant clonotypes only (FDR < 0.05) — note the bimodal pattern"
  )
```

## V Gene Enrichment in Expanders

Do significantly expanding clonotypes preferentially use certain TRBV gene segments?

```{r vgene-enrichment, fig.height=5}
# Get V gene info for significant clonotypes
expander_cdr3 <- all_results %>%
  filter(direction == "Expander") %>%
  select(patient_id, cdr3_aa)

expander_vgenes <- tcr_data %>%
  filter(sample_type == "Product") %>%
  inner_join(expander_cdr3, by = c("patient_id", "cdr3_aa")) %>%
  count(v_gene, name = "expander_count")

# Background V gene usage in Product
background_vgenes <- tcr_data %>%
  filter(sample_type == "Product") %>%
  count(v_gene, name = "total_count")

vgene_enrichment <- expander_vgenes %>%
  left_join(background_vgenes, by = "v_gene") %>%
  mutate(
    expander_frac  = expander_count / sum(expander_count),
    background_frac = total_count / sum(total_count),
    enrichment = log2(expander_frac / background_frac)
  ) %>%
  arrange(desc(enrichment)) %>%
  head(15)

ggplot(vgene_enrichment, aes(x = reorder(v_gene, enrichment), y = enrichment,
                              fill = enrichment > 0)) +
  geom_col(alpha = 0.8, width = 0.6) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#00ff9d", "FALSE" = "#ff6b6b"), guide = "none") +
  geom_hline(yintercept = 0, color = "#7b9db5", linewidth = 0.5) +
  labs(
    x = NULL, y = expression(log[2] ~ "Enrichment (Expanders vs Background)"),
    title = "V Gene Enrichment in Significant Expanders",
    subtitle = "Positive values indicate over-representation among expanding clonotypes"
  )
```

## Summary

> **Key Findings from Differential Abundance Analysis:**
>
> 1. **edgeR identifies hundreds of significantly differentially abundant clonotypes** per patient (FDR < 0.05), with the majority being contractors — consistent with the massive diversity reduction observed during manufacturing.
> 2. **Significant expanders represent the clonotypes selected and amplified** during the manufacturing process. These are the candidate therapeutic clones.
> 3. **CR patients show distinct expansion patterns** compared to PD patients, with the most dramatic fold changes in the expander population.
> 4. **Volcano plots reveal strong asymmetry** — the contractor population is larger but the expander population shows more extreme fold changes, indicating focused amplification of select clones.
> 5. **V gene enrichment analysis** suggests potential biases in which TCR gene segments are preferentially expanded during manufacturing.

### Statistical Framework Notes

- **Model:** edgeR negative binomial GLM with common + tagwise dispersion
- **Test:** Likelihood ratio test (Product vs Apheresis)
- **Multiple testing:** Benjamini-Hochberg FDR correction
- **Significance threshold:** FDR < 0.05
- **Note:** With only 2 samples per patient (no biological replicates), dispersion estimates are approximate. In a real clinical setting, biological replicates or a multi-patient GLM design would provide more robust inference.
