---
title: "Clinical Response Integration"
subtitle: "TCR Repertoire Analysis — Notebook 06"
author: "Joshua Luthy"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    css: "../www/report-style.css"
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 6,
  dev = "png", dpi = 150, fig.retina = 2
)
library(tidyverse)

theme_tcr <- function(base_size = 13) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.background    = element_rect(fill = "#0a0f14", color = NA),
      panel.background   = element_rect(fill = "#0a0f14", color = NA),
      panel.grid.major   = element_line(color = "#1e3348", linewidth = 0.3),
      panel.grid.minor   = element_blank(),
      text               = element_text(color = "#e0eef5", family = "sans"),
      axis.text          = element_text(color = "#7b9db5", size = 10),
      axis.title         = element_text(color = "#7b9db5", size = 11),
      plot.title         = element_text(color = "#e0eef5", face = "bold", size = 16,
                                        margin = margin(b = 8)),
      plot.subtitle      = element_text(color = "#7b9db5", size = 11,
                                        margin = margin(b = 12)),
      legend.background  = element_rect(fill = "#0f1923", color = "#1e3348"),
      legend.text        = element_text(color = "#7b9db5"),
      legend.title       = element_text(color = "#e0eef5"),
      strip.text         = element_text(color = "#e0eef5", face = "bold"),
      plot.margin        = margin(15, 15, 15, 15)
    )
}

theme_set(theme_tcr())
```

## Analysis Overview

This final notebook **integrates findings across the full pipeline** to build a
comprehensive picture of how TCR repertoire features relate to clinical response.
We combine diversity metrics, clonal expansion data, AUC kinetics, and edgeR
differential abundance results into a unified patient-level feature matrix and
explore correlations with treatment outcome (CR, PR, PD).

```{r load-all-data}
# Load all processed datasets
diversity   <- read_csv("../data/processed/diversity_metrics.csv")
auc_metrics <- read_csv("../data/processed/clonal_auc_metrics.csv")
top_clones  <- read_csv("../data/processed/top_clonotypes.csv")
patient_meta <- read_csv("../data/raw/patient_metadata.csv")
longitudinal <- read_csv("../data/processed/longitudinal_clonal_tracking.csv")
```

## Building the Feature Matrix

We aggregate all per-patient metrics into a single feature matrix for clinical
correlation analysis.

```{r build-features}
# Diversity features (Product only — characterizes the drug product)
product_diversity <- diversity %>%
  filter(sample_type == "Product") %>%
  select(patient_id, clinical_response,
         product_shannon    = shannon_entropy,
         product_clonality  = clonality,
         product_richness   = n_clonotypes,
         product_d50        = d50)

# Diversity fold change (Apheresis → Product)
diversity_fc <- diversity %>%
  select(patient_id, sample_type, shannon_entropy, n_clonotypes) %>%
  pivot_wider(names_from = sample_type, values_from = c(shannon_entropy, n_clonotypes)) %>%
  mutate(
    shannon_fc    = shannon_entropy_Product / shannon_entropy_Apheresis,
    richness_fc   = n_clonotypes_Product / n_clonotypes_Apheresis
  ) %>%
  select(patient_id, shannon_fc, richness_fc)

# AUC features (aggregated per patient)
auc_features <- auc_metrics %>%
  group_by(patient_id, clinical_response) %>%
  summarise(
    mean_auc          = mean(auc),
    max_auc           = max(auc),
    mean_persistence  = mean(persistence_ratio),
    mean_peak_freq    = mean(peak_frequency),
    n_persistent      = sum(persistence_ratio > 0.5),
    .groups = "drop"
  ) %>%
  select(-clinical_response)

# Top clone dominance
top1_features <- top_clones %>%
  filter(sample_type == "Product", rank == 1) %>%
  select(patient_id, top1_fraction = clone_fraction)

# Combine all features
feature_matrix <- product_diversity %>%
  left_join(diversity_fc, by = "patient_id") %>%
  left_join(auc_features, by = "patient_id") %>%
  left_join(top1_features, by = "patient_id") %>%
  left_join(patient_meta %>% select(patient_id, age, sex, prior_lines), by = "patient_id")

feature_matrix %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  knitr::kable(caption = "Integrated feature matrix — one row per patient")
```

## Correlation Heatmap

We examine pairwise correlations between all numeric features to identify the
strongest predictors of clinical response grouping.

```{r correlation-heatmap, fig.height=7, fig.width=8}
# Compute correlation matrix
numeric_features <- feature_matrix %>%
  select(where(is.numeric), -age, -prior_lines) %>%
  select(-product_d50)  # Remove integer-only column for cleaner correlations

cor_mat <- cor(numeric_features, use = "pairwise.complete.obs", method = "spearman")

# Reshape for ggplot heatmap
cor_long <- as.data.frame(as.table(cor_mat)) %>%
  rename(feature1 = Var1, feature2 = Var2, correlation = Freq)

ggplot(cor_long, aes(x = feature1, y = feature2, fill = correlation)) +
  geom_tile(color = "#1e3348", linewidth = 0.5) +
  geom_text(aes(label = round(correlation, 2)),
            color = ifelse(abs(cor_long$correlation) > 0.6, "#0a0f14", "#7b9db5"),
            size = 3) +
  scale_fill_gradient2(
    low = "#ff6b6b", mid = "#0a0f14", high = "#00ff9d",
    midpoint = 0, limits = c(-1, 1)
  ) +
  labs(
    title = "Feature Correlation Matrix (Spearman)",
    subtitle = "Green = positive correlation, Red = negative",
    fill = "ρ"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    axis.title   = element_blank()
  )
```

## Feature Distributions by Response

### Key Metrics Across Response Groups

```{r feature-panels, fig.height=12, fig.width=10}
# Reshape for faceted plotting
plot_features <- feature_matrix %>%
  select(patient_id, clinical_response,
         `Product Shannon H`   = product_shannon,
         `Product Clonality`   = product_clonality,
         `Mean AUC`            = mean_auc,
         `Mean Persistence`    = mean_persistence,
         `Top Clone Fraction`  = top1_fraction,
         `Shannon FC (P/A)`    = shannon_fc) %>%
  pivot_longer(-c(patient_id, clinical_response),
               names_to = "metric", values_to = "value")

ggplot(plot_features, aes(x = clinical_response, y = value,
                           fill = clinical_response, color = clinical_response)) +
  geom_col(data = . %>% group_by(clinical_response, metric) %>%
             summarise(value = mean(value), .groups = "drop"),
           alpha = 0.2, width = 0.5) +
  geom_point(size = 4, alpha = 0.8) +
  scale_fill_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  scale_color_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +
  labs(
    x = "Clinical Response", y = "Value",
    title = "Key Features Stratified by Clinical Response",
    subtitle = "Individual patient values overlaid on group means"
  ) +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "#152231", color = "#1e3348"),
    strip.text = element_text(color = "#00d4ff", face = "bold", size = 10)
  )
```

## Composite Score

We construct a simple **composite score** that summarizes the key features associated
with clinical response. Features are z-score normalized and combined with direction-aware
weighting (higher clonality, AUC, and persistence are favorable; higher Shannon entropy
in Product is unfavorable).

```{r composite-score, fig.height=5}
# Z-score normalization
scored <- feature_matrix %>%
  mutate(
    z_clonality   = scale(product_clonality)[,1],
    z_auc         = scale(mean_auc)[,1],
    z_persistence = scale(mean_persistence)[,1],
    z_top1        = scale(top1_fraction)[,1],
    z_shannon_inv = -scale(product_shannon)[,1],  # invert: lower diversity = higher score
    
    # Composite: equal weight
    composite_score = (z_clonality + z_auc + z_persistence + z_top1 + z_shannon_inv) / 5
  )

ggplot(scored, aes(x = reorder(patient_id, -composite_score), y = composite_score,
                    fill = clinical_response)) +
  geom_col(alpha = 0.8, width = 0.6) +
  geom_text(aes(label = round(composite_score, 2)),
            vjust = ifelse(scored$composite_score > 0, -0.5, 1.5),
            color = "#e0eef5", size = 4, fontface = "bold") +
  scale_fill_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  geom_hline(yintercept = 0, color = "#3f6380", linewidth = 0.5) +
  labs(
    x = "Patient", y = "Composite Score",
    fill = "Response",
    title = "Composite Repertoire Score by Patient",
    subtitle = "Higher scores reflect more focused Product, stronger expansion, and greater persistence"
  )
```

> **Interpretation:** The composite score separates patients by clinical response,
> with CR patients scoring highest (most focused Product, strongest expansion, best
> persistence) and PD patients scoring lowest.

## Radar / Spider Plot — Patient Profiles

```{r radar-data, fig.height=8, fig.width=10}
# Normalize each feature to 0-1 range for radar visualization
radar_data <- scored %>%
  select(patient_id, clinical_response,
         Clonality = product_clonality,
         `Top Clone %` = top1_fraction,
         `Mean AUC` = mean_auc,
         Persistence = mean_persistence,
         `N Persistent` = n_persistent) %>%
  mutate(across(Clonality:`N Persistent`, ~ (. - min(.)) / (max(.) - min(.) + 1e-10)))

# Reshape for parallel coordinate plot (approximation of radar)
radar_long <- radar_data %>%
  pivot_longer(-c(patient_id, clinical_response),
               names_to = "metric", values_to = "value") %>%
  mutate(metric = factor(metric, levels = c("Clonality", "Top Clone %", "Mean AUC",
                                             "Persistence", "N Persistent")))

ggplot(radar_long, aes(x = metric, y = value, group = patient_id,
                        color = clinical_response)) +
  geom_line(linewidth = 1.1, alpha = 0.8) +
  geom_point(size = 3) +
  scale_color_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  coord_polar() +
  labs(
    title = "Patient Repertoire Profiles",
    subtitle = "Normalized metrics — CR patients (green) show consistently higher values",
    color = "Response"
  ) +
  theme(
    axis.text.y  = element_blank(),
    axis.title   = element_blank(),
    panel.grid.major = element_line(color = "#1e3348", linewidth = 0.3)
  )
```

## Statistical Testing — Response Group Comparisons

```{r kruskal-tests}
# Kruskal-Wallis tests across response groups for key features
test_features <- c("product_shannon", "product_clonality", "mean_auc",
                    "mean_persistence", "top1_fraction")

kw_results <- map_dfr(test_features, function(feat) {
  formula <- as.formula(paste(feat, "~ clinical_response"))
  test <- kruskal.test(formula, data = feature_matrix)
  tibble(
    feature   = feat,
    statistic = round(test$statistic, 3),
    p_value   = round(test$p.value, 4),
    significant = test$p.value < 0.05
  )
})

kw_results %>%
  knitr::kable(caption = "Kruskal-Wallis tests: feature ~ clinical response")
```

> **Note:** With n=2 per group, statistical power is extremely limited and p-values
> should be interpreted as descriptive trends rather than confirmatory evidence. In a
> real clinical trial, these analyses would be powered with larger cohorts.

## Summary Heatmap — Patient × Feature

```{r summary-heatmap, fig.height=5, fig.width=9}
# Z-scored heatmap of all key features
heatmap_data <- scored %>%
  select(patient_id, clinical_response,
         z_clonality, z_auc, z_persistence, z_top1, z_shannon_inv) %>%
  pivot_longer(-c(patient_id, clinical_response),
               names_to = "feature", values_to = "z_score") %>%
  mutate(feature = recode(feature,
    z_clonality   = "Product Clonality",
    z_auc         = "Mean AUC",
    z_persistence = "Persistence",
    z_top1        = "Top Clone Dominance",
    z_shannon_inv = "Product Focus (inv. Shannon)"
  ))

ggplot(heatmap_data, aes(x = feature, y = reorder(patient_id, -z_score), fill = z_score)) +
  geom_tile(color = "#1e3348", linewidth = 0.8) +
  geom_text(aes(label = round(z_score, 1)),
            color = ifelse(abs(heatmap_data$z_score) > 0.8, "#0a0f14", "#e0eef5"),
            size = 3.5, fontface = "bold") +
  scale_fill_gradient2(
    low = "#ff6b6b", mid = "#152231", high = "#00ff9d",
    midpoint = 0, limits = c(-2, 2)
  ) +
  labs(
    x = NULL, y = NULL, fill = "Z-Score",
    title = "Patient × Feature Heatmap (Z-Scored)",
    subtitle = "Green = favorable (higher expansion/persistence), Red = unfavorable"
  ) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 9),
    axis.text.y = element_text(size = 11)
  )
```

## Conclusions

> **Integrated findings across the full TCR analysis pipeline:**
>
> 1. **Product repertoire focusing predicts response.** CR patients have the lowest Product
>    Shannon entropy, highest clonality, and most dominant top clones — manufacturing efficiently
>    selected a narrow set of therapeutic clonotypes.
>
> 2. **Post-infusion expansion kinetics separate responders from non-responders.** CR patients
>    show stronger, later-peaking, and more persistent clonal expansion (higher AUC, higher
>    persistence ratio).
>
> 3. **The composite score integrating diversity, expansion, and persistence separates
>    clinical response groups**, suggesting that a multi-metric approach may be more informative
>    than any single measure.
>
> 4. **These patterns are consistent across multiple analytical approaches** — descriptive
>    diversity metrics (Notebook 02), clonal dominance analysis (Notebook 03), edgeR
>    differential abundance (Notebook 04), longitudinal tracking (Notebook 05), and integrated
>    feature analysis (this notebook) all converge on the same biological conclusion.

### Limitations & Future Directions

- **Sample size (n=6)** limits statistical power; larger cohorts needed for confirmatory analysis
- **Synthetic data** — patterns are designed to recapitulate known biology but are not derived from real patient samples
- **Predictive modeling** (logistic regression, random forest) would require larger cohorts to train and validate
- **Antigen specificity** analysis (e.g., linking expanded TCRs to known epitopes via databases like VDJdb or IEDB) would add biological interpretability
- **Multi-omic integration** (scRNA-seq, cytokine data, clinical labs) could further refine the response prediction model
