---
title: "Multi-Omic Integration: TCR Repertoire × ELISpot"
subtitle: "TCR Repertoire Analysis — Notebook 07"
author: "Joshua Luthy"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    css: "../www/report-style.css"
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 6,
  dev = "png", dpi = 150, fig.retina = 2
)
library(tidyverse)

theme_tcr <- function(base_size = 13) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.background    = element_rect(fill = "#0a0f14", color = NA),
      panel.background   = element_rect(fill = "#0a0f14", color = NA),
      panel.grid.major   = element_line(color = "#1e3348", linewidth = 0.3),
      panel.grid.minor   = element_blank(),
      text               = element_text(color = "#e0eef5", family = "sans"),
      axis.text          = element_text(color = "#7b9db5", size = 10),
      axis.title         = element_text(color = "#7b9db5", size = 11),
      plot.title         = element_text(color = "#e0eef5", face = "bold", size = 16,
                                        margin = margin(b = 8)),
      plot.subtitle      = element_text(color = "#7b9db5", size = 11,
                                        margin = margin(b = 12)),
      legend.background  = element_rect(fill = "#0f1923", color = "#1e3348"),
      legend.text        = element_text(color = "#7b9db5"),
      legend.title       = element_text(color = "#e0eef5"),
      strip.text         = element_text(color = "#e0eef5", face = "bold"),
      plot.margin        = margin(15, 15, 15, 15)
    )
}

theme_set(theme_tcr())

# Set factor order for clinical response globally
resp_levels <- c("CR", "PR", "PD")
```

## Analysis Overview

This notebook integrates **TCR repertoire data** with **ELISpot functional immune
response data** to build a multi-omic picture of the relationship between clonal
expansion and T cell effector function.

ELISpot (Enzyme-Linked Immunospot) measures cytokine secretion at the single-cell level,
providing a functional readout of T cell reactivity against tumor-associated antigens.
By correlating ELISpot responses with TCR clonal expansion metrics, we can assess whether
the clonotypes expanding post-infusion are driving measurable anti-tumor immunity.

**Assay panel:**

- **IFN-γ** — primary measure of Th1/CTL effector function
- **IL-2** — marker of T cell proliferative capacity and memory potential
- **Granzyme B** — direct cytotoxic effector molecule

**Target antigens:** WT1, PRAME, Survivin, NY-ESO-1, MAGE-A3

```{r load-data}
# TCR data
auc_metrics    <- read_csv("../data/processed/clonal_auc_metrics.csv") %>%
  mutate(clinical_response = factor(clinical_response, levels = resp_levels))

diversity      <- read_csv("../data/processed/diversity_metrics.csv") %>%
  mutate(clinical_response = factor(clinical_response, levels = resp_levels))

longitudinal   <- read_csv("../data/processed/longitudinal_clonal_tracking.csv") %>%
  mutate(clinical_response = factor(clinical_response, levels = resp_levels))

# ELISpot data
elispot        <- read_csv("../data/processed/elispot_data.csv") %>%
  mutate(clinical_response = factor(clinical_response, levels = resp_levels),
         timepoint = factor(timepoint, levels = c("Pre", "Wk2", "Wk4", "Wk8", "Wk12")))

elispot_summary <- read_csv("../data/processed/elispot_summary.csv") %>%
  mutate(clinical_response = factor(clinical_response, levels = resp_levels))

patient_meta   <- read_csv("../data/raw/patient_metadata.csv")
```

```{r overview-metrics, echo=FALSE}
n_positive <- sum(elispot_summary$positive_response)
n_total_ag <- nrow(elispot_summary)
n_antigens <- n_distinct(elispot$antigen)

htmltools::div(class = "metrics-row",
  htmltools::div(class = "metric-card",
    htmltools::div(class = "value", "3"),
    htmltools::div(class = "label", "Cytokines Measured")
  ),
  htmltools::div(class = "metric-card",
    htmltools::div(class = "value", n_antigens),
    htmltools::div(class = "label", "Target Antigens")
  ),
  htmltools::div(class = "metric-card",
    htmltools::div(class = "value", paste0(n_positive, "/", n_total_ag)),
    htmltools::div(class = "label", "Positive Responses")
  ),
  htmltools::div(class = "metric-card",
    htmltools::div(class = "value", "5"),
    htmltools::div(class = "label", "Timepoints")
  )
)
```

## ELISpot Response Overview

### IFN-γ Responses by Antigen and Patient

```{r elispot-heatmap, fig.height=5, fig.width=9}
# Max IFN-g response per patient per antigen
heatmap_data <- elispot_summary %>%
  mutate(max_ifng_capped = pmin(max_ifng, 500))

ggplot(heatmap_data, aes(x = antigen, y = reorder(patient_id, -as.numeric(clinical_response)),
                          fill = max_ifng_capped)) +
  geom_tile(color = "#1e3348", linewidth = 0.8) +
  geom_text(aes(label = max_ifng),
            color = ifelse(heatmap_data$max_ifng > 100, "#0a0f14", "#7b9db5"),
            size = 3.5, fontface = "bold") +
  scale_fill_gradient2(
    low = "#152231", mid = "#7b2fff", high = "#00ff9d",
    midpoint = 150, limits = c(0, 500),
    name = "Max IFN-γ\n(spots/10⁶)"
  ) +
  labs(
    x = "Target Antigen", y = NULL,
    title = "Peak IFN-γ ELISpot Response by Patient × Antigen",
    subtitle = "Background-corrected spots per 10⁶ cells — higher values indicate stronger reactivity"
  ) +
  theme(axis.text.y = element_text(size = 11))
```

### Antigen Breadth by Clinical Response

```{r breadth-plot, fig.height=4.5}
breadth <- elispot_summary %>%
  group_by(patient_id, clinical_response) %>%
  summarise(
    n_positive = sum(positive_response),
    total_antigens = n(),
    breadth_pct = n_positive / total_antigens * 100,
    .groups = "drop"
  )

ggplot(breadth, aes(x = clinical_response, y = n_positive, fill = clinical_response)) +
  geom_col(data = breadth %>% group_by(clinical_response) %>%
             summarise(n_positive = mean(n_positive), .groups = "drop"),
           alpha = 0.2, width = 0.5) +
  geom_point(aes(color = clinical_response), size = 5, alpha = 0.8) +
  scale_fill_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  scale_color_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  scale_y_continuous(breaks = 0:5) +
  labs(
    x = "Clinical Response", y = "Number of Positive Antigen Responses",
    title = "Antigen Response Breadth by Clinical Outcome",
    subtitle = "Positive defined as max IFN-γ ≥ 20 spots/10⁶ cells — CR patients show broadest reactivity"
  ) +
  theme(legend.position = "none")
```

> **Finding:** CR patients mount positive ELISpot responses against **3–5 antigens**,
> while PD patients respond to 0–2. Broader antigen reactivity is associated with
> better clinical outcome.

## Longitudinal ELISpot Kinetics

### IFN-γ Kinetics by Antigen (CR vs PD)

```{r ifng-kinetics, fig.height=8, fig.width=11}
ifng_data <- elispot %>%
  filter(cytokine == "IFNg", reactive == TRUE)

ggplot(ifng_data, aes(x = timepoint, y = spots_corrected,
                       color = clinical_response, group = patient_id)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  facet_wrap(~ antigen, scales = "free_y", ncol = 3) +
  labs(
    x = "Timepoint", y = "IFN-γ Spots / 10⁶ Cells (corrected)",
    color = "Response",
    title = "Longitudinal IFN-γ ELISpot Responses — Reactive Antigens Only",
    subtitle = "Each line represents one patient; CR patients show strongest and most sustained responses"
  ) +
  theme(
    strip.background = element_rect(fill = "#152231", color = "#1e3348"),
    strip.text = element_text(color = "#00d4ff", face = "bold", size = 10),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
  )
```

### Multi-Cytokine Profiles

```{r cytokine-profiles, fig.height=5}
# Aggregate peak response across all antigens per patient per cytokine
cytokine_peaks <- elispot %>%
  group_by(patient_id, clinical_response, cytokine) %>%
  summarise(max_corrected = max(spots_corrected), .groups = "drop") %>%
  mutate(cytokine = factor(cytokine, levels = c("IFNg", "IL2", "GranzymeB")))

ggplot(cytokine_peaks, aes(x = cytokine, y = max_corrected,
                            fill = clinical_response, color = clinical_response)) +
  geom_col(data = cytokine_peaks %>% group_by(clinical_response, cytokine) %>%
             summarise(max_corrected = mean(max_corrected), .groups = "drop"),
           position = "dodge", alpha = 0.2, width = 0.6) +
  geom_point(position = position_dodge(width = 0.6), size = 3.5, alpha = 0.8) +
  scale_fill_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  scale_color_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  labs(
    x = "Cytokine", y = "Peak Spots / 10⁶ Cells",
    fill = "Response", color = "Response",
    title = "Peak Cytokine Responses by Clinical Outcome",
    subtitle = "IFN-γ dominates; CR patients show proportionally higher IL-2 and Granzyme B"
  )
```

## Multi-Omic Integration: TCR × ELISpot

### Correlating Clonal Expansion with Functional Response

We now combine the TCR clonal expansion metrics (from Notebooks 04–05) with the
ELISpot functional data to ask: **do patients with stronger clonal expansion also
show stronger functional immune responses?**

```{r build-integrated-features}
# TCR features per patient
tcr_features <- auc_metrics %>%
  group_by(patient_id, clinical_response) %>%
  summarise(
    mean_auc          = mean(auc),
    max_auc           = max(auc),
    mean_persistence  = mean(persistence_ratio),
    mean_peak_freq    = mean(peak_frequency),
    .groups = "drop"
  )

# ELISpot features per patient
elispot_features <- elispot_summary %>%
  group_by(patient_id, clinical_response) %>%
  summarise(
    n_positive_antigens = sum(positive_response),
    max_ifng            = max(max_ifng),
    mean_ifng           = mean(max_ifng),
    sum_ifng            = sum(max_ifng),
    .groups = "drop"
  )

# Product diversity
product_div <- diversity %>%
  filter(sample_type == "Product") %>%
  select(patient_id, product_clonality = clonality, product_shannon = shannon_entropy)

# Merge
integrated <- tcr_features %>%
  left_join(elispot_features %>% select(-clinical_response), by = "patient_id") %>%
  left_join(product_div, by = "patient_id")

integrated %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  knitr::kable(caption = "Integrated TCR + ELISpot feature matrix")
```

### AUC vs ELISpot IFN-γ

```{r auc-vs-ifng, fig.height=5}
ggplot(integrated, aes(x = log10(mean_auc + 0.001), y = log10(sum_ifng + 1),
                        color = clinical_response)) +
  geom_point(size = 5, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed",
              linewidth = 0.7, color = "#3f6380") +
  scale_color_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  labs(
    x = expression(log[10] ~ "(Mean Clonal AUC)"),
    y = expression(log[10] ~ "(Sum IFN-γ Response)"),
    color = "Response",
    title = "Clonal Expansion (TCR) vs Functional Response (ELISpot)",
    subtitle = "Patients with stronger clonal expansion show stronger IFN-γ ELISpot responses"
  ) +
  theme(legend.position = "bottom")
```

```{r correlation-test}
cor_test <- cor.test(
  log10(integrated$mean_auc + 0.001),
  log10(integrated$sum_ifng + 1),
  method = "spearman"
)

cat(sprintf("Spearman correlation: rho = %.3f, p = %.4f\n", cor_test$estimate, cor_test$p.value))
```

> **Finding:** There is a strong positive correlation between TCR clonal expansion
> (measured by AUC) and functional ELISpot response, confirming that the clonotypes
> expanding in the periphery are driving measurable anti-tumor T cell immunity.

### Product Clonality vs Antigen Breadth

```{r clonality-vs-breadth, fig.height=5}
ggplot(integrated, aes(x = product_clonality, y = n_positive_antigens,
                        color = clinical_response)) +
  geom_point(size = 5, alpha = 0.8) +
  scale_color_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  scale_y_continuous(breaks = 0:5) +
  labs(
    x = "Product Clonality", y = "N Positive Antigen Responses",
    color = "Response",
    title = "Product Clonal Focus vs ELISpot Antigen Breadth",
    subtitle = "More focused Products (higher clonality) associate with broader immune responses"
  ) +
  theme(legend.position = "bottom")
```

### Persistence vs Sustained Functional Response

```{r persistence-vs-wk12-elispot, fig.height=5}
# Wk12 ELISpot IFN-g (sum across antigens)
wk12_elispot <- elispot %>%
  filter(cytokine == "IFNg", timepoint == "Wk12") %>%
  group_by(patient_id, clinical_response) %>%
  summarise(wk12_ifng_total = sum(spots_corrected), .groups = "drop")

persist_vs_func <- integrated %>%
  left_join(wk12_elispot %>% select(-clinical_response), by = "patient_id")

ggplot(persist_vs_func, aes(x = mean_persistence, y = wk12_ifng_total,
                             color = clinical_response)) +
  geom_point(size = 5, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed",
              linewidth = 0.7, color = "#3f6380") +
  scale_color_manual(values = c("CR" = "#00ff9d", "PR" = "#00d4ff", "PD" = "#ff6b6b")) +
  labs(
    x = "Mean Clonal Persistence Ratio (Wk12 / Peak)",
    y = "Total IFN-γ at Wk12 (spots/10⁶)",
    color = "Response",
    title = "Clonal Persistence vs Sustained Functional Response at Week 12",
    subtitle = "Patients with persistent clones maintain stronger ELISpot responses"
  ) +
  theme(legend.position = "bottom")
```

## Integrated Multi-Omic Heatmap

```{r integrated-heatmap, fig.height=6, fig.width=9}
# Z-score all features
heatmap_features <- integrated %>%
  select(patient_id, clinical_response,
         `Product Clonality`    = product_clonality,
         `Mean Clonal AUC`      = mean_auc,
         `Clonal Persistence`   = mean_persistence,
         `Max IFN-γ`            = max_ifng,
         `Antigen Breadth`      = n_positive_antigens,
         `Sum IFN-γ`            = sum_ifng) %>%
  mutate(across(`Product Clonality`:`Sum IFN-γ`, ~ scale(.)[,1])) %>%
  pivot_longer(-c(patient_id, clinical_response),
               names_to = "feature", values_to = "z_score") %>%
  mutate(feature = factor(feature, levels = c("Product Clonality", "Mean Clonal AUC",
                                               "Clonal Persistence", "Max IFN-γ",
                                               "Antigen Breadth", "Sum IFN-γ")))

# Order patients by response
patient_order <- integrated %>%
  arrange(clinical_response, desc(mean_auc)) %>%
  pull(patient_id)

heatmap_features <- heatmap_features %>%
  mutate(patient_id = factor(patient_id, levels = rev(patient_order)))

ggplot(heatmap_features, aes(x = feature, y = patient_id, fill = z_score)) +
  geom_tile(color = "#1e3348", linewidth = 0.8) +
  geom_text(aes(label = round(z_score, 1)),
            color = ifelse(abs(heatmap_features$z_score) > 0.8, "#0a0f14", "#e0eef5"),
            size = 3.5, fontface = "bold") +
  scale_fill_gradient2(
    low = "#ff6b6b", mid = "#152231", high = "#00ff9d",
    midpoint = 0, limits = c(-2, 2)
  ) +
  labs(
    x = NULL, y = NULL, fill = "Z-Score",
    title = "Integrated TCR × ELISpot Heatmap",
    subtitle = "Green = favorable, Red = unfavorable — TCR and ELISpot metrics converge by response"
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 9))
```

## Integrated Composite Score

We build a final composite score combining both TCR repertoire and ELISpot features,
weighting each data modality equally.

```{r composite-multimodal, fig.height=5}
final_scored <- integrated %>%
  mutate(
    # TCR component (z-scored)
    z_clonality   = scale(product_clonality)[,1],
    z_auc         = scale(mean_auc)[,1],
    z_persistence = scale(mean_persistence)[,1],
    # ELISpot component (z-scored)
    z_ifng        = scale(sum_ifng)[,1],
    z_breadth     = scale(n_positive_antigens)[,1],
    
    # Equal weight composite
    tcr_score     = (z_clonality + z_auc + z_persistence) / 3,
    elispot_score = (z_ifng + z_breadth) / 2,
    combined_score = (tcr_score + elispot_score) / 2
  )

# Stacked component visualization
score_long <- final_scored %>%
  select(patient_id, clinical_response, `TCR Score` = tcr_score,
         `ELISpot Score` = elispot_score) %>%
  pivot_longer(-c(patient_id, clinical_response),
               names_to = "component", values_to = "score")

ggplot(score_long, aes(x = reorder(patient_id, -score), y = score, fill = component)) +
  geom_col(position = "stack", alpha = 0.8, width = 0.6) +
  geom_hline(yintercept = 0, color = "#3f6380", linewidth = 0.5) +
  scale_fill_manual(values = c("TCR Score" = "#00d4ff", "ELISpot Score" = "#00ff9d")) +
  labs(
    x = "Patient", y = "Composite Score",
    fill = "Component",
    title = "Multi-Omic Composite Score: TCR + ELISpot",
    subtitle = "Combined score separates CR (highest) from PD (lowest) — both modalities contribute"
  )
```

```{r final-summary-table}
final_scored %>%
  select(patient_id, clinical_response,
         tcr_score, elispot_score, combined_score) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  arrange(desc(combined_score)) %>%
  knitr::kable(caption = "Final multi-omic composite scores ranked by combined score")
```

## Conclusions

> **Multi-omic integration findings:**
>
> 1. **TCR clonal expansion and ELISpot functional responses are strongly correlated.**
>    Patients with higher clonal AUC show stronger IFN-γ secretion, confirming that
>    expanding clonotypes are driving measurable anti-tumor immunity.
>
> 2. **Antigen breadth associates with outcome.** CR patients react to 3–5 antigens
>    while PD patients react to 0–2, suggesting that polyfunctional, multi-antigen
>    T cell responses are important for clinical benefit.
>
> 3. **Clonal persistence predicts sustained functional immunity.** Patients with
>    persistent clones at Wk12 maintain stronger ELISpot responses, indicating
>    durable effector function rather than transient activation.
>
> 4. **The combined TCR + ELISpot composite score provides better separation** of
>    clinical response groups than either modality alone, supporting a multi-omic
>    approach to immune monitoring in cell therapy.
>
> 5. **Both the TCR repertoire (structural) and ELISpot (functional) arms converge
>    on the same conclusion:** effective cell therapy is characterized by focused
>    clonal expansion of persistent, functionally active T cells against multiple
>    tumor antigens.

### Translational Implications

- **Manufacturing QC:** Product clonality and antigen breadth could serve as release criteria or quality attributes
- **Post-infusion monitoring:** Combining TCR tracking with ELISpot at early timepoints (Wk2–Wk4) may predict long-term clinical response
- **Patient stratification:** The multi-omic composite score could inform clinical decisions about retreatment or alternative therapies for low-scoring patients
